---
title: "10.4 Report Exercise"
author: "Manuel Summermatter"
date: "2023-05-22"
output: html_document
---

First, I set up the work space
```{r}
library(tidyverse)
library(lubridate)
library(caret)
library(recipes)
source(paste0(here::here(), "/R/eval_model.R"))
source(paste0(here::here(), "/R/eval_model_chapter_10.R"))
```

Now, I load and prepare the data
```{r}
fluxes_Davos <- read_csv(paste0(here::here(), "/data/FLX_CH-Dav_FLUXNET2015_FULLSET_DD_1997-2014_1-3.csv")) |>  
  dplyr::select(TIMESTAMP,
                GPP_NT_VUT_REF,
                ends_with("_QC"),
                ends_with("_F"),
                -contains("JSB")
                ) |>
  dplyr::mutate(TIMESTAMP = ymd(TIMESTAMP)) |>
  dplyr::mutate(across(where(is.numeric), ~na_if(., -9999))) |>
  dplyr::mutate(GPP_NT_VUT_REF = ifelse(NEE_VUT_REF_QC < 0.8, NA, GPP_NT_VUT_REF),
                TA_F           = ifelse(TA_F_QC        < 0.8, NA, TA_F),
                SW_IN_F        = ifelse(SW_IN_F_QC     < 0.8, NA, SW_IN_F),
                LW_IN_F        = ifelse(LW_IN_F_QC     < 0.8, NA, LW_IN_F),
                VPD_F          = ifelse(VPD_F_QC       < 0.8, NA, VPD_F),
                PA_F           = ifelse(PA_F_QC        < 0.8, NA, PA_F),
                P_F            = ifelse(P_F_QC         < 0.8, NA, P_F),
                WS_F           = ifelse(WS_F_QC        < 0.8, NA, WS_F)) |>
  dplyr::select(-ends_with("_QC"))


fluxes_Laegern <- read_csv(paste0(here::here(), "/data/FLX_CH-Lae_FLUXNET2015_FULLSET_DD_2004-2014_1-4.csv")) |>  
  dplyr::select(TIMESTAMP,
                GPP_NT_VUT_REF,
                ends_with("_QC"),
                ends_with("_F"),
                -contains("JSB"),
                -P_F
                ) |>
  dplyr::mutate(TIMESTAMP = ymd(TIMESTAMP)) |>
  dplyr::mutate(across(where(is.numeric), ~na_if(., -9999))) |>
  dplyr::mutate(GPP_NT_VUT_REF = ifelse(NEE_VUT_REF_QC < 0.8, NA, GPP_NT_VUT_REF),
                TA_F           = ifelse(TA_F_QC        < 0.8, NA, TA_F),
                SW_IN_F        = ifelse(SW_IN_F_QC     < 0.8, NA, SW_IN_F),
                LW_IN_F        = ifelse(LW_IN_F_QC     < 0.8, NA, LW_IN_F),
                VPD_F          = ifelse(VPD_F_QC       < 0.8, NA, VPD_F),
                PA_F           = ifelse(PA_F_QC        < 0.8, NA, PA_F),
                WS_F           = ifelse(WS_F_QC        < 0.8, NA, WS_F)) |>
  dplyr::select(-ends_with("_QC"))
```

Now, I get an overview of the data
```{r}
gridExtra::grid.arrange(
  fluxes_Davos |> 
    ggplot(aes(x = GPP_NT_VUT_REF, y = ..count..)) + 
    geom_histogram() +
    labs(title = "Davos"),
  fluxes_Laegern |> 
    ggplot(aes(x = GPP_NT_VUT_REF, y = ..count..)) + 
    geom_histogram() +
    labs(title = "Laegern")
)
```

What follows are further steps
```{r}
# Data splitting Davos
set.seed(1982)
split_Davos <- rsample::initial_split(fluxes_Davos, prop = 0.8, strata = "VPD_F")
fluxes_Davos_train <- rsample::training(split_Davos)
fluxes_Davos_test <- rsample::testing(split_Davos)

# Data splitting Laegern
set.seed(1982)
split_Laegern <- rsample::initial_split(fluxes_Laegern, prop = 0.8, strata = "VPD_F")
fluxes_Laegern_train <- rsample::training(split_Laegern)
fluxes_Laegern_test <- rsample::testing(split_Laegern)


# Model and pre-processing formulation Davos, use all variables but LW_IN_F
pp_Davos <- recipes::recipe(GPP_NT_VUT_REF ~ SW_IN_F + VPD_F + TA_F, 
                      data = fluxes_Davos_train |> drop_na()) |> 
  recipes::step_BoxCox(all_predictors()) |> 
  recipes::step_center(all_numeric(), -all_outcomes()) |>
  recipes::step_scale(all_numeric(), -all_outcomes())

# Model and pre-processing formulation Laegern, use all variables but LW_IN_F
pp_Laegern <- recipes::recipe(GPP_NT_VUT_REF ~ SW_IN_F + VPD_F + TA_F, 
                      data = fluxes_Laegern_train |> drop_na()) |> 
  recipes::step_BoxCox(all_predictors()) |> 
  recipes::step_center(all_numeric(), -all_outcomes()) |>
  recipes::step_scale(all_numeric(), -all_outcomes())


# KNN model for Davos
mod_knn_Davos <- caret::train(
  pp_Davos, 
  data = fluxes_Davos_train |> drop_na(), 
  method = "knn",
  trControl = caret::trainControl(method = "none"),
  tuneGrid = data.frame(k = 8),
  metric = "RMSE"
)

# KNN model for Laegern
mod_knn_Laegern <- caret::train(
  pp_Laegern, 
  data = fluxes_Laegern_train |> drop_na(), 
  method = "knn",
  trControl = caret::trainControl(method = "none"),
  tuneGrid = data.frame(k = 8),
  metric = "RMSE"
)

# P_F variable in fluxes_Laegern_train has NA in every row, so after the drop_na() it drops every row and the dataframe is empty -> Then obviously the model can't work with this empty dataframe
```

Now, I call the model evaluation function for the KNN model for Davos and Laegern, each tested within-site and across-site
```{r}
# KNN prediction for Davos, within-site
eval_model_chapter_10(mod = mod_knn_Davos, df_train = fluxes_Davos_train, df_test = fluxes_Davos_test)

# KNN prediction for Davos, across-site
eval_model_chapter_10(mod = mod_knn_Davos, df_train = fluxes_Davos_train, df_test = fluxes_Laegern_test)

# KNN prediction for Laegern, within-site
eval_model_chapter_10(mod = mod_knn_Laegern, df_train = fluxes_Laegern_train, df_test = fluxes_Laegern_test)

# KNN prediction for Laegern, across-site
eval_model_chapter_10(mod = mod_knn_Laegern, df_train = fluxes_Laegern_train, df_test = fluxes_Davos_test)
```

The KNN model which is trained on the Davos site performs better when predicted on the within-site test set than on the across-site test set:  
- R squared is 0.64 within-site and 0.48 across-site  
- RMSE is 1.56 within-site and 3.52 across-site  
- MAE is within-site 1.17 and 2.78 across-site  

For the knn model which is trained on the Laegern site, the results point to the same findings. The model performs better on the within-site prediction than on the across-site prediction:  
- R squared is 0.66 within-site and 0.49 across-site  
- RMSE is 2.52 within-site and 2.64 across-site  
- MAE is within-site 1.91 and 2.08 across-site