---
title: "3.5 Report Exercise"
author: "Manuel Summermatter"
date: "2023-03-11"
output: html_document
---
# Cleaning data from elevated CO2 experiments

## 2.
First, I set up the work space and load the data which I have manually cleaned in excel
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
elevated_co2_experiments <- read_delim("C:/Manuel Summermatter/Bachelorstudium/6. Semester/Applied Geodata Science I/agds_report_ManuelSummermatter/data/Elevated_CO2_Experiments.csv", col_types = "cccc?iiiicc", skip = 3)
head(elevated_co2_experiments)
```

Now, I add a new column to the dataframe: The log-response ratio (RR)
```{r}
elevated_co2_experiments <-  elevated_co2_experiments |>
  mutate(RR = log(SOC_elevated / SOC_ambient))
head(elevated_co2_experiments)
```

Lastly, I aggregate the data per experiment and calculate the log-response ratio within each experiment
```{r}  
RR_within_each_experiment <- elevated_co2_experiments|>
  group_by(Experiment) |>
  summarise(RR = mean(RR))
knitr::kable(RR_within_each_experiment)
```

## 3.
I start the third part of this exercise with work space cleaning, removing the most recently created dataframe "RR_within_each_experiment"
```{r}
rm(RR_within_each_experiment)
```

Now, I divide the dataframe into 3 different ones, distinguished by their respective year since the start of the experiment: The early, mid- and late phase. Lastly, I put these dataframes back together, so that it results in one table again
```{r}
early_phase <- elevated_co2_experiments |>
  filter(Time_years < 3) |>
  mutate(phase = "early")

mid_phase <- elevated_co2_experiments |>
  filter(Time_years >= 3 & Time_years <= 6) |>
  mutate(phase = "mid")

late_phase <- elevated_co2_experiments |>
  filter(Time_years > 6) |>
  mutate(phase = "late")

end_table <- bind_rows(early_phase, mid_phase, late_phase)
head(end_table)
```

Now, I create a table which represents the log-response ratio for each of the three phases
```{r}
result_table <- end_table |>
  group_by(phase) |>
  summarise(RR = mean(RR, na.rm = TRUE))
knitr::kable(result_table)
```